---
title: "Analysis"
output:
  html_document:
    toc: true
    toc_float: true
---

```{r include = FALSE}
library(tidyverse)
library(dplyr)
library(broom)
library(knitr)
library(kableExtra)
library("here")
```

```{r}
source(here("data", "datacleaning.R"))
```


### Introduction

This analysis aims to investigate factors associated with the likelihood of being diagnosed with diabetes (DIBEV_A) using a multivariable logistic regression model. Covariates include demographic, lifestyle, physical health, and chronic disease indicators.

### Methodology

The outcome variable, DIBEV_A, is coded as a binary factor (0 = No diabetes, 1 = Diabetes diagnosis). Several categorical predictors are converted to factor type, and continuous variables such as age, height, and weight are included directly. A logistic regression model with a logit link is fitted.

The original model : 

$$
\text{Logit}(P(\text{DIBEV_A}=1)) = \log\left(\frac{P(\text{DIBEV_A}=1)}{1-P(\text{DIBEV_A}=1)}\right) = \beta_0 \\
+ \beta_1 \cdot \text{SEX_A} + \beta_2 \cdot \text{AGEP_A} + \beta_3 \cdot \text{BMICATD_A} + \beta_4 \cdot \text{WEIGHTLBTC_A} + \beta_5 \cdot \text{HEIGHTTC_A} \\
+ \beta_6 \cdot \text{DIFF_A} + \beta_7 \cdot \text{EDUCP_A} + \beta_8 \cdot \text{LSATIS4_A} + \beta_9 \cdot \text{PHSTAT_A} + \beta_{10} \cdot \text{CHLEV_A} \\
+ \beta_{11} \cdot \text{HYPEV_A} + \beta_{12} \cdot \text{LIVEREV_A} + \beta_{13} \cdot \text{STREV_A} + \beta_{14} \cdot \text{MIEV_A} + \beta_{15} \cdot \text{ANGEV_A} \\
+ \beta_{16} \cdot \text{CHDEV_A} + \beta_{17} \cdot \text{DEPEV_A} + \beta_{18} \cdot \text{ANXEV_A} + \beta_{19} \cdot \text{SMKEV_A} + \beta_{20} \cdot \text{LONELY_A}
$$

### Build Model


```{r}
df_model = 
  nhis24_df_fit |>
  mutate(
    # Outcome: DIBEV_A
    DIBEV_A = factor(DIBEV_A, levels = c(0, 1)),
    
    # Covariates - Dichotomous/Ordinal/Categorical (MUST be factor)
    SEX_A = factor(SEX_A), # Assume 1/2 coding is acceptable
    BMICATD_A = factor(BMICATD_A), # Codes 0, 1, 2, 3
    DIFF_A = factor(DIFF_A), # Codes 0, 1, 2, 3
    EDUCP_A = factor(EDUCP_A), # Codes 0, 1, 2
    LSATIS4_A = factor(LSATIS4_A), # Codes 0, 1
    PHSTAT_A = factor(PHSTAT_A), # Codes 0, 1
    CHLEV_A = factor(CHLEV_A), # Codes 0, 1
    HYPEV_A = factor(HYPEV_A), # Codes 0, 1
    LIVEREV_A = factor(LIVEREV_A), # Codes 0, 1
    STREV_A = factor(STREV_A), # Codes 0, 1
    MIEV_A = factor(MIEV_A), # Codes 0, 1
    ANGEV_A = factor(ANGEV_A), # Codes 0, 1
    CHDEV_A = factor(CHDEV_A), # Codes 0, 1
    DEPEV_A = factor(DEPEV_A), # Codes 0, 1
    ANXEV_A = factor(ANXEV_A), # Codes 0, 1
    SMKEV_A = factor(SMKEV_A), # Codes 0, 1
    LONELY_A = factor(LONELY_A) # Codes 0, 1, 2, 3, 4
  ) 
```

```{r}
model_formula <- DIBEV_A ~ SEX_A + AGEP_A + BMICATD_A + WEIGHTLBTC_A + HEIGHTTC_A + 
                           DIFF_A + EDUCP_A + LSATIS4_A + PHSTAT_A + CHLEV_A + 
                           HYPEV_A + LIVEREV_A + STREV_A + MIEV_A + ANGEV_A + 
                           CHDEV_A + DEPEV_A + ANXEV_A + SMKEV_A + LONELY_A

glm_model <- glm(
  model_formula, 
  data = df_model,
  family = binomial(link = "logit")
)

summary(glm_model)
```

### Model Results

```{r}
tidy_results = 
  glm_model |>
  tidy(conf.int = TRUE, exponentiate = TRUE) |>
   rename(
     OR = estimate,
     CI_low = conf.low,
     CI_high = conf.high
)

kable(tidy_results, digits = 4, caption = "Logistic Regression Model Results") |>
kable_styling(full_width = FALSE)

```

