---
title: "Analysis"
output:
  html_document:
    code_folding: hide
    toc: true
    toc_float: true
---

```{r include = FALSE}
library(tidyverse)
library(dplyr)
library(broom)
library(knitr)
library(kableExtra)
library("here")
library(ggplot2)
```

```{r}
source(here("data", "datacleaning.R"))
```


# Introduction

This analysis aims to investigate factors associated with the likelihood of being diagnosed with diabetes (DIBEV_A) using a multivariable logistic regression model. Covariates include demographic, lifestyle, physical health, and chronic disease indicators.

# Methodology

The outcome variable, `DIBEV_A`, is a binary indicator representing whether an individual has been diagnosed with diabetes (0 = No diabetes, 1 = Diabetes diagnosis). Given the binary nature of the outcome, a logistic regression model within the Generalized Linear Model (GLM) framework was employed. Logistic regression is appropriate for modeling the probability of an event occurring while ensuring predicted probabilities remain between 0 and 1, using a logit link to relate the linear combination of predictors to the log-odds of the outcome.

Covariates in the model include demographic factors (e.g., sex and age), anthropometric measures (weight and height), lifestyle indicators, and a range of chronic disease or health status variables. Categorical variables, including BMI category, educational attainment, and various health event indicators, were encoded as factors to allow estimation of separate effects for each non-reference level. Continuous variables such as age, weight, and height were included directly. 

The logistic regression model can be formally expressed as:

$$
\text{Logit}(P(\text{DIBEV_A}=1)) = \beta_0 
+ \sum_{i=1}^{20} \beta_i X_i
$$

where $X_i$ represents each covariate, including both categorical indicator variables and continuous predictors, and $\beta_i$ denotes the corresponding log-odds coefficient. Exponentiating the coefficients yields odds ratios, which provide a clinically interpretable measure of the relative likelihood of diabetes associated with each predictor, conditional on all other variables in the model.


The outcome variable, DIBEV_A, is coded as a binary factor (0 = No diabetes, 1 = Diabetes diagnosis). Several categorical predictors are converted to factor type, and continuous variables such as age, height, and weight are included directly. A logistic regression model with a logit link is fitted.

The original detailed model : 

$$
\text{Logit}(P(\text{DIBEV_A}=1)) = \log\left(\frac{P(\text{DIBEV_A}=1)}{1-P(\text{DIBEV_A}=1)}\right) = \beta_0 \\
+ \beta_1 \cdot \text{SEX_A} + \beta_2 \cdot \text{AGEP_A} + \beta_3 \cdot \text{BMICATD_A} + \beta_4 \cdot \text{WEIGHTLBTC_A} + \beta_5 \cdot \text{HEIGHTTC_A} \\
+ \beta_6 \cdot \text{DIFF_A} + \beta_7 \cdot \text{EDUCP_A} + \beta_8 \cdot \text{LSATIS4_A} + \beta_9 \cdot \text{PHSTAT_A} + \beta_{10} \cdot \text{CHLEV_A} \\
+ \beta_{11} \cdot \text{HYPEV_A} + \beta_{12} \cdot \text{LIVEREV_A} + \beta_{13} \cdot \text{STREV_A} + \beta_{14} \cdot \text{MIEV_A} + \beta_{15} \cdot \text{ANGEV_A} \\
+ \beta_{16} \cdot \text{CHDEV_A} + \beta_{17} \cdot \text{DEPEV_A} + \beta_{18} \cdot \text{ANXEV_A} + \beta_{19} \cdot \text{SMKEV_A} + \beta_{20} \cdot \text{LONELY_A}
$$

# Build Model

```{r}
df_model = 
  nhis24_df_fit |>
  mutate(
    # Outcome: DIBEV_A
    DIBEV_A = factor(DIBEV_A, levels = c(0, 1)),
    
    # Covariates - Dichotomous/Ordinal/Categorical (MUST be factor)
    SEX_A = factor(SEX_A), # Assume 1/2 coding is acceptable
    BMICATD_A = factor(BMICATD_A), # Codes 0, 1, 2, 3
    DIFF_A = factor(DIFF_A), # Codes 0, 1, 2, 3
    EDUCP_A = factor(EDUCP_A), # Codes 0, 1, 2
    LSATIS4_A = factor(LSATIS4_A), # Codes 0, 1
    PHSTAT_A = factor(PHSTAT_A), # Codes 0, 1
    CHLEV_A = factor(CHLEV_A), # Codes 0, 1
    HYPEV_A = factor(HYPEV_A), # Codes 0, 1
    LIVEREV_A = factor(LIVEREV_A), # Codes 0, 1
    STREV_A = factor(STREV_A), # Codes 0, 1
    MIEV_A = factor(MIEV_A), # Codes 0, 1
    ANGEV_A = factor(ANGEV_A), # Codes 0, 1
    CHDEV_A = factor(CHDEV_A), # Codes 0, 1
    DEPEV_A = factor(DEPEV_A), # Codes 0, 1
    ANXEV_A = factor(ANXEV_A), # Codes 0, 1
    SMKEV_A = factor(SMKEV_A), # Codes 0, 1
    LONELY_A = factor(LONELY_A) # Codes 0, 1, 2, 3, 4
  ) 
```

```{r}
model_formula <- DIBEV_A ~ SEX_A + AGEP_A + BMICATD_A + WEIGHTLBTC_A + HEIGHTTC_A + 
                           DIFF_A + EDUCP_A + LSATIS4_A + PHSTAT_A + CHLEV_A + 
                           HYPEV_A + LIVEREV_A + STREV_A + MIEV_A + ANGEV_A + 
                           CHDEV_A + DEPEV_A + ANXEV_A + SMKEV_A + LONELY_A

glm_model <- glm(
  model_formula, 
  data = df_model,
  family = binomial(link = "logit")
)

summary(glm_model)
```

## histogram of P-value
```{r}
glm_summary = summary(glm_model)

p_values_df = data.frame(
  Variable = rownames(glm_summary$coefficients),
  P_value = glm_summary$coefficients[, "Pr(>|z|)"]
)

# Remove intercept
p_values_df = 
  p_values_df[p_values_df$Variable != "(Intercept)", ]

P_value_hisg =
  p_values_df|>
  ggplot(aes(x = reorder(Variable, P_value), y = P_value)) +
  geom_bar(
    stat = "identity",
    fill = ifelse(p_values_df$P_value < 0.05, "red", "skyblue")
  ) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "black") +
  coord_flip() +
  labs(
    title = "P-values from Logistic Regression",
    x = "Variables",
    y = "P-value"
  ) +
  theme_minimal()

P_value_hisg
```

# Model Results

```{r}
model_results_tidy = 
  tidy(glm_model, conf.int = TRUE) |>
  mutate(
    Beta = estimate,
    OR = exp(estimate),
    Lower_CI = exp(conf.low),
    Upper_CI = exp(conf.high),
    
    # Clean up term names for presentation
    term = gsub("`", "", term), 
    term = case_when(
        term == "(Intercept)" ~ "Intercept",
        term == "SEX_A2" ~ "SEX_A: Female (vs Male)",
        
        # BMI (0: Healthy Weight)
        term == "BMICATD_A1" ~ "BMICATD_A: Underweight (vs Healthy)",
        term == "BMICATD_A2" ~ "BMICATD_A: Overweight (vs Healthy)",
        term == "BMICATD_A3" ~ "BMICATD_A: Obese (vs Healthy)",
        
        # Difficulty Walking (0: No Difficulty)
        term == "DIFF_A1" ~ "DIFF_A: Some Difficulty (vs No Difficulty)",
        term == "DIFF_A2" ~ "DIFF_A: A lot of Difficulty (vs No Difficulty)",
        term == "DIFF_A3" ~ "DIFF_A: Cannot do (vs No Difficulty)",
        
        # Education (0: Low Education)
        term == "EDUCP_A1" ~ "EDUCP_A: Medium Edu (vs Low Edu)",
        term == "EDUCP_A2" ~ "EDUCP_A: High Edu (vs Low Edu)",
        
        # Life Satisfaction (0: Satisfied/Very Satisfied)
        term == "LSATIS4_A1" ~ "LSATIS4_A: Dissatisfied (vs Satisfied)",
        
        # Health Status (0: Good/VGood/Excellent Health)
        term == "PHSTAT_A1" ~ "PHSTAT_A: Fair/Poor Health (vs Good Health)",
        
        # Loneliness (0: Never Lonely)
        term == "LONELY_A1" ~ "LONELY_A: Rarely Lonely (vs Never Lonely)",
        term == "LONELY_A2" ~ "LONELY_A: Sometimes Lonely (vs Never Lonely)",
        term == "LONELY_A3" ~ "LONELY_A: Usually Lonely (vs Never Lonely)",
        term == "LONELY_A4" ~ "LONELY_A: Always Lonely (vs Never Lonely)",
        
        # Chronic Dichotomous Conditions (1: Yes vs 0: No)
        term == "CHLEV_A1" ~ "CHLEV_A: High Cholesterol (Yes)",
        term == "HYPEV_A1" ~ "HYPEV_A: Hypertension (Yes)",
        term == "LIVEREV_A1" ~ "LIVEREV_A: Liver Disease (Yes)",
        term == "STREV_A1" ~ "STREV_A: Stroke (Yes)",
        term == "MIEV_A1" ~ "MIEV_A: Myocardial Infarction (Yes)",
        term == "ANGEV_A1" ~ "ANGEV_A: Angina (Yes)",
        term == "CHDEV_A1" ~ "CHDEV_A: CHD (Yes)",
        term == "DEPEV_A1" ~ "DEPEV_A: Depression (Yes)",
        term == "ANXEV_A1" ~ "ANXEV_A: Anxiety (Yes)",
        term == "SMKEV_A1" ~ "SMKEV_A: Smoker (Yes)",
        
        # Continuous Variables
        term == "AGEP_A" ~ "AGEP_A (Per Year)",
        term == "WEIGHTLBTC_A" ~ "WEIGHTLBTC_A (Per lb)",
        term == "HEIGHTTC_A" ~ "HEIGHTTC_A (Per unit)",
        TRUE ~ term
    ),
    p.value = format.pval(p.value, digits = 3, eps = 0.001)
  ) |>
  select(Term = term, Beta, OR, Lower_CI, Upper_CI, p.value)

# Generate kable table
model_results_tidy |>
  mutate(
    OR_CI = paste0(round(OR, 3), " (", round(Lower_CI, 3), " - ", round(Upper_CI, 3), ")"),
    Beta = round(Beta, 4)
  ) |>
  select(Term, Beta, OR_CI, p.value) |>
  kable(
    caption = "Multivariable Logistic Regression of Diabetes Diagnosis (Adjusted Odds Ratios)",
    col.names = c("Predictor Term", "Coefficient (Beta)", "Adjusted OR (95% CI)", "P-value"),
    align = c('l', 'r', 'r', 'r'),
    booktabs = TRUE
  ) |>
  kable_styling(
    full_width = TRUE, 
    position = "left",
    latex_options = "hold_position"
  ) |>
  add_header_above(c(" " = 2, "Odds Ratio" = 1, " " = 1)) %>%
  footnote(
    general = "Reference Outcome: No Diabetes (DIBEV_A = 0).",
    threeparttable = TRUE
  )
```

# Forest Results
First of all, we filter out terms that are very close to 1 with wide CIs and non-significant to reduce clutter. Secondly, we order the terms for better visualization, e.g., by OR value.
```{r}
forest_data = 
  model_results_tidy |>
  filter(Term != "Intercept") |>
  mutate(Term = fct_reorder(Term, OR, .desc = TRUE)) 

forest_plots = 
ggplot(forest_data, aes(x = OR, y = Term)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red") +
  geom_pointrange(aes(xmin = Lower_CI, xmax = Upper_CI), size = 0.3) +
  scale_x_log10(breaks = c(0.1, 0.2, 0.5, 1, 2, 5, 10),
                labels = c("0.1", "0.2", "0.5", "1", "2", "5", "10")) + 
  labs(
    title = "Forest Plot of Adjusted Odds Ratios for Diabetes Diagnosis",
    x = "Adjusted Odds Ratio (log scale)",
    y = "Predictor Term"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.y = element_text(size = 9, face = "bold"),
    axis.title.x = element_text(size = 10, face = "bold"),
    axis.title.y = element_text(size = 10, face = "bold"),
    panel.grid.minor.x = element_blank() # Remove minor grid lines on x-axis
  )

print(forest_plots)
```

This forest plot displays adjusted odds ratios (ORs) for diabetes diagnosis, where the y-axis lists predictors (with reference groups in parentheses), and the x-axis (log scale) shows ORs (dashed red line at OR=1, no association). Based on the plots, we used the 95% confidence interval (CI) exclusion of 1 as the measure of statistical significance, several variables demonstrated a significant independent association with a diabetes diagnosis. Predictors with OR < 1 and 95% CIs not crossing 1 (e.g., underweight BMI, higher education, lower loneliness, smoking, anxiety) are linked to lower diabetes risk; those with OR > 1 and significant CIs (e.g., overweight/obese BMI, liver disease, fair/poor health, high cholesterol, hypertension) increase risk; and factors like stroke history, severe mobility difficulty, and some loneliness categories have CIs crossing 1, showing no significant association.



# Interpretation of Model Coefficients

Below is the interpretation of the logistic regression output, with each variable summarized under its own subsection.

---

### Sex
#### Male vs Female
- Coefficient β = –0.4402  
- OR = 0.644 (95% CI: 0.602–0.688), *p* < 0.001  

Males had significantly lower odds of diabetes compared with females after adjusting for all covariates.

---

### Age
#### Continuous (per year increase)
- OR = 1.002 (95% CI: 1.000–1.003), *p* = 0.0156  

Older age was a small but statistically significant risk factor for diabetes.

---

### Smoking
#### Smoker vs Non-Smoker
- OR = 0.916 (95% CI: 0.838–1.001), *p* = 0.0534  

Smoking showed a borderline non-significant inverse association.  
There is no strong evidence of an association between smoking and diabetes in this adjusted model.

---

### Loneliness
Compared with the “Never Lonely” reference group:

- Rarely lonely: OR = 1.042, *p* = 0.4543  
- Sometimes lonely: OR = 1.071, *p* = 0.2363  
- Usually lonely: OR = 0.835, *p* = 0.2029  
- Always lonely: OR = 1.162, *p* = 0.2244  

None of the loneliness categories were statistically significant.  
Loneliness does not appear to be independently associated with diabetes risk.

---

### Education
#### Medium Education vs Low Education
- OR = 0.896 (95% CI: 0.809–0.993), *p* = 0.0364  

#### High Education vs Low Education
- OR = 0.725 (95% CI: 0.651–0.807), *p* < 0.001  

Higher education was protective.  
Those with high education had markedly lower odds of diabetes.

---

### BMI Category
Reference = BMI category 0 (Normal weight)

- Underweight: OR = 1.887, *p* < 0.001  
- Overweight: OR = 2.314, *p* < 0.001  
- Obesity (Class I–III): OR = 4.491, *p* < 0.001  

BMI demonstrated a strong dose-response relationship, with obesity being the strongest risk factor in the model.

---

### Difficulty Walking (Mobility)
Reference = No difficulty

- Some difficulty: OR = 1.173, *p* = 0.0037  
- A lot of difficulty: OR = 1.373, *p* < 0.001  
- Cannot do at all: OR = 1.273, *p* = 0.0831  

Greater mobility limitations were associated with higher odds of diabetes.

---

### Life Satisfaction
- Dissatisfied vs satisfied: OR = 0.884, *p* = 0.2083 → Not significant  

---

### General Health Status
#### Fair/Poor vs Good/Very Good/Excellent
- OR = 2.004 (95% CI: 1.805–2.224), *p* < 0.001  

Poor health was a strong predictor of diabetes.

---

### High Cholesterol
- OR = 2.107, *p* < 0.001  

A strong and significant association.

---

### Hypertension
- OR = 2.128, *p* < 0.001  

Hypertension was one of the largest predictors, more than doubling the odds of diabetes.

---

### Liver Disease
- OR = 1.928, *p* < 0.001  

Individuals with liver disease had nearly double the odds of diabetes.

---

### Stroke
- OR = 1.112, *p* = 0.199  

Not statistically significant.

---

### Myocardial Infarction
- OR = 1.329 (95% CI: 1.118–1.577), *p* = 0.0012  

History of myocardial infarction was positively associated with diabetes.

---

### Angina
- OR = 1.008, *p* = 0.942  

No association.

---

### Coronary Heart Disease (CHD)
- OR = 1.258 (95% CI: 1.089–1.451), *p* = 0.0018  

CHD was a significant predictor of diabetes.

---

### Depression
- OR = 1.175 (95% CI: 1.035–1.333), *p* = 0.0122  

Depression was significantly associated with increased diabetes odds.

---

### Anxiety
- OR = 0.921 (95% CI: 0.808–1.049), *p* = 0.216  

Anxiety was not significantly related to diabetes.

---

### Summary of Findings
Overall, the multivariable logistic regression analysis identifies several strong predictors of diabetes. Obesity, hypertension, high cholesterol, liver disease, poor general health, and myocardial infarction were among the strongest risk factors, each contributing substantially to elevated diabetes odds. Higher education showed a consistent protective effect. Mobility limitations were also associated with greater risk, illustrating the interconnected nature of physical functioning and metabolic health. While depression was significantly associated with increased diabetes risk, other psychosocial factors such as anxiety and loneliness were not significant in this adjusted model. Smoking likewise showed no significant association. Collectively, these findings highlight the multifactorial nature of diabetes, involving biological, behavioral, and socioeconomic determinants.



