---
title: "Analysis"
---

```{r include = FALSE}
library(tidyverse)
library(dplyr)
library(broom)
library(knitr)
library(kableExtra)
library("here")
library(ggplot2)
```

```{r include = FALSE}
knitr::opts_chunk$set(
  fig.width = 9,
  fig.asp = .6,
  out.width = "100%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

```{r}
source(here("data", "datacleaning.R"))
```

  
# Introduction

This analysis aims to investigate factors associated with the likelihood of being diagnosed with diabetes (DIBEV_A) using a multivariable logistic regression model. Covariates include demographic, lifestyle, physical health, and chronic disease indicators.
  
# Methodology

The outcome variable, `DIBEV_A`, is a binary indicator representing whether an individual has been diagnosed with diabetes (0 = No diabetes, 1 = Diabetes diagnosis). Given the binary nature of the outcome, a logistic regression model within the Generalized Linear Model (GLM) framework was employed. Logistic regression is appropriate for modeling the probability of an event occurring while ensuring predicted probabilities remain between 0 and 1, using a logit link to relate the linear combination of predictors to the log-odds of the outcome.

Covariates in the model include demographic factors (e.g., sex and age), anthropometric measures (weight and height), lifestyle indicators, and a range of chronic disease or health status variables. Categorical variables, including BMI category, educational attainment, and various health event indicators, were encoded as factors to allow estimation of separate effects for each non-reference level. Continuous variables such as age, weight, and height were included directly. 

The logistic regression model can be formally expressed as:

$$
\text{Logit}(P(\text{DIBEV_A}=1)) = \beta_0 
+ \sum_{i=1}^{20} \beta_i X_i
$$

where $X_i$ represents each covariate, including both categorical indicator variables and continuous predictors, and $\beta_i$ denotes the corresponding log-odds coefficient. Exponentiating the coefficients yields odds ratios, which provide a clinically interpretable measure of the relative likelihood of diabetes associated with each predictor, conditional on all other variables in the model.


The outcome variable, DIBEV_A, is coded as a binary factor (0 = No diabetes, 1 = Diabetes diagnosis). Several categorical predictors are converted to factor type, and continuous variables such as age, height, and weight are included directly. A logistic regression model with a logit link is fitted.

# Correlation Analysis

To better understand the relationships between the variables in our dataset, we find the need to conduct a correlation analysis. The original dataset was obtained from the 2024 NHIS (National Health Interview Survey), which contains information on demographic characteristics, health conditions, and behavioral factors for a nationally representative sample of adults. The cleaned dataset included variables such as age, sex, weight, height, and various health indicators.  

***Correlation analysis*** serves as a preliminary step to identify variables that are linearly associated. Understanding these relationships helps us:
-	Detect potential multicollinearity before performing regression analyses or more complex modeling.
-	Provide evidence-based justification for including or excluding certain variables in our following statistical models.

We computed ***pairwise Pearson correlations*** for all numeric variables. Correlations with an absolute value greater than 0.3 are considered moderate or strong associations. Additionally, diagonal elements (self-correlations) were excluded from interpretation, and symmetric duplicates were removed to ensure clarity.  

To understand ***Coefficient of determination*** ($R^2$), $R^2$ represents the proportion of the variance in one variable that is predictable from the other variable.
For our selected threshold of $r = 0.3$:
$$R^2 = (0.3)^2 = 0.09$$
This means that for two variables correlated at $r = 0.3$, 9% of the variation in one variable is explained by the other variable. 

```{r}
# Select numeric variables
nhis24_df_fit_corr = nhis24_df_fit |> 
  select(-BMI_A)

numeric_vars <- nhis24_df_fit_corr[, sapply(nhis24_df_fit_corr, is.numeric)]

# Compute correlation matrix
cor_matrix <- cor(numeric_vars, use = "pairwise.complete.obs")

# Show matrix
print(cor_matrix)

cor_pt = as.data.frame(as.table(cor_matrix)) |> 
  ggplot(aes(Var1, Var2, fill = Freq)) +
  geom_tile() +
  scale_fill_gradient2(
    low = "blue", mid = "white", high = "red",
    midpoint = 0, 
    name = "Correlation"  
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

cor_pt
```

Key Findings:
From our analysis, the following moderate-to-strong correlations were observed:
```{r}
# > 0.03 in cor_matrix
# Convert correlation matrix to a data frame
cor_df <- as.data.frame(as.table(cor_matrix))

# Rename columns
colnames(cor_df) <- c("Var1", "Var2", "Correlation")
# Filter |correlation| > 0.3 but exclude self-correlations
high_cor <- cor_df |> 
  filter(abs(Correlation) > 0.3 & Var1 != Var2)

# remove duplicate
high_cor_unique <- high_cor %>%
  rowwise() %>%
  mutate(pair = paste(sort(c(Var1, Var2)), collapse = "_")) %>%
  distinct(pair, .keep_all = TRUE) %>%
  select(-pair) |> 
  arrange(desc(Correlation))

knitr::kable(high_cor_unique, caption = "Variables with Correlation > 0.3")
```

By choosing $r > 0.3$, we are strategically focusing our analysis on variables that exhibit a relationship deemed moderate or stronger, filtering out very weak or negligible associations that may not be meaningful for our research or modeling goals.

As highlighted in the correlation analysis above, height and weight exhibited a strong linear association, an observation that raised concerns about potential multicollinearity in subsequent regression modeling. Multicollinearity can distort coefficient estimates and reduce model reliability, so we opted to exclude these two highly correlated variables from the model. In their place, we introduced a continuous BMI variable using the raw height and weight data. This replacement strategy offers dual advantages: it eliminates the multicollinearity risk posed by height and weight while integrating their combined biological information into a single, interpretable metric, thereby adequately addressing confounding from these two factors.

Below is the specification of the initial full logistic regression model (prior to further variable selection):

$$
\text{Logit}(P(\text{DIBEV_A}=1)) = \log\left(\frac{P(\text{DIBEV_A}=1)}{1-P(\text{DIBEV_A}=1)}\right) = \beta_0 \\
+ \beta_1 \cdot \text{SEX_A} + \beta_2 \cdot \text{AGEP_A} + \beta_3 \cdot \text{BMI_A} + \beta_4 \cdot \text{DIFF_A} + \beta_5 \cdot \text{EDUCP_A} \\
+ \beta_6 \cdot \text{LSATIS4_A} + \beta_7 \cdot \text{PHSTAT_A} + \beta_8 \cdot \text{CHLEV_A} + \beta_9 \cdot \text{HYPEV_A} + \beta_{10} \cdot \text{LIVEREV_A} \\
+ \beta_{11} \cdot \text{STREV_A} + \beta_{12} \cdot \text{MIEV_A} + \beta_{13} \cdot \text{ANGEV_A} + \beta_{14} \cdot \text{CHDEV_A} + \beta_{15} \cdot \text{DEPEV_A} \\
+ \beta_{16} \cdot \text{ANXEV_A} + \beta_{17} \cdot \text{SMKEV_A} + \beta_{18} \cdot \text{LONELY_A}
$$

# Build Model

```{r}
df_model = 
  nhis24_df_fit |>
  mutate(
    # Outcome: DIBEV_A
    DIBEV_A = factor(DIBEV_A, levels = c(0, 1)),
    
    # Covariates - Dichotomous/Ordinal/Categorical (MUST be factor)
    SEX_A = factor(SEX_A), # Assume 1/2 coding is acceptable
    DIFF_A = factor(DIFF_A), # Codes 0, 1, 2, 3
    EDUCP_A = factor(EDUCP_A), # Codes 0, 1, 2
    LSATIS4_A = factor(LSATIS4_A), # Codes 0, 1
    PHSTAT_A = factor(PHSTAT_A), # Codes 0, 1
    CHLEV_A = factor(CHLEV_A), # Codes 0, 1
    HYPEV_A = factor(HYPEV_A), # Codes 0, 1
    LIVEREV_A = factor(LIVEREV_A), # Codes 0, 1
    STREV_A = factor(STREV_A), # Codes 0, 1
    MIEV_A = factor(MIEV_A), # Codes 0, 1
    ANGEV_A = factor(ANGEV_A), # Codes 0, 1
    CHDEV_A = factor(CHDEV_A), # Codes 0, 1
    DEPEV_A = factor(DEPEV_A), # Codes 0, 1
    ANXEV_A = factor(ANXEV_A), # Codes 0, 1
    SMKEV_A = factor(SMKEV_A), # Codes 0, 1
    LONELY_A = factor(LONELY_A) # Codes 0, 1, 2, 3, 4
  ) 
```

```{r}
model_formula <- DIBEV_A ~ SEX_A + AGEP_A + BMI_A +
                           DIFF_A + EDUCP_A + LSATIS4_A + PHSTAT_A + CHLEV_A + 
                           HYPEV_A + LIVEREV_A + STREV_A + MIEV_A + ANGEV_A + 
                           CHDEV_A + DEPEV_A + ANXEV_A + SMKEV_A + LONELY_A

glm_model <- glm(
  model_formula, 
  data = df_model,
  family = binomial(link = "logit")
)

summary(glm_model)
```

## Bar chart of P-value
```{r}
glm_summary = summary(glm_model)

p_values_df = data.frame(
  Variable = rownames(glm_summary$coefficients),
  P_value = glm_summary$coefficients[, "Pr(>|z|)"]
)

# Remove intercept
p_values_df = 
  p_values_df[p_values_df$Variable != "(Intercept)", ]

P_value_hisg =
  p_values_df|>
  ggplot(aes(x = reorder(Variable, P_value), y = P_value)) +
  geom_bar(
    stat = "identity",
    fill = ifelse(p_values_df$P_value < 0.05, "red", "skyblue")
  ) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "black") +
  coord_flip() +
  scale_y_continuous(limits = c(0, 0.5)) +
  labs(
    title = "P-values from Logistic Regression",
    x = "Variables",
    y = "P-value"
  ) +
  theme_minimal()

P_value_hisg
```
  
# Model Results

Tidy results.

```{r}
tidy_logit <- function(model) {
  broom::tidy(model, conf.int = TRUE) |>
    mutate(
      Beta     = estimate,
      OR       = exp(estimate),
      Lower_CI = exp(conf.low),
      Upper_CI = exp(conf.high),
      p.value  = format.pval(p.value, digits = 3, eps = 0.001)
    ) |>
    select(term, Beta, OR, Lower_CI, Upper_CI, p.value)
}
```

Clean up term names for presentation.

```{r}
clean_term_labels <- function(df) {
  df |>
    mutate(
      term = gsub("`", "", term),
      Term = case_when(
        term == "(Intercept)" ~ "Intercept",
        term == "SEX_A2" ~ "SEX_A: Female (vs Male)",
        
        # Difficulty Walking (0: No Difficulty)
        term == "DIFF_A1" ~ "DIFF_A: Some Difficulty (vs No Difficulty)",
        term == "DIFF_A2" ~ "DIFF_A: A lot of Difficulty (vs No Difficulty)",
        term == "DIFF_A3" ~ "DIFF_A: Cannot do (vs No Difficulty)",
        
        # Education (0: Low Education)
        term == "EDUCP_A1" ~ "EDUCP_A: Medium Edu (vs Low Edu)",
        term == "EDUCP_A2" ~ "EDUCP_A: High Edu (vs Low Edu)",
        
        # Life Satisfaction (0: Satisfied/Very Satisfied)
        term == "LSATIS4_A1" ~ "LSATIS4_A: Dissatisfied (vs Satisfied)",
        
        # Health Status (0: Good/VGood/Excellent Health)
        term == "PHSTAT_A1" ~ "PHSTAT_A: Fair/Poor Health (vs Good Health)",
        
        # Loneliness (0: Never Lonely)
        term == "LONELY_A1" ~ "LONELY_A: Rarely Lonely (vs Never Lonely)",
        term == "LONELY_A2" ~ "LONELY_A: Sometimes Lonely (vs Never Lonely)",
        term == "LONELY_A3" ~ "LONELY_A: Usually Lonely (vs Never Lonely)",
        term == "LONELY_A4" ~ "LONELY_A: Always Lonely (vs Never Lonely)",
        
        # Chronic Dichotomous Conditions (1: Yes vs 0: No)
        term == "CHLEV_A1" ~ "CHLEV_A: High Cholesterol (Yes)",
        term == "HYPEV_A1" ~ "HYPEV_A: Hypertension (Yes)",
        term == "LIVEREV_A1" ~ "LIVEREV_A: Liver Disease (Yes)",
        term == "STREV_A1" ~ "STREV_A: Stroke (Yes)",
        term == "MIEV_A1" ~ "MIEV_A: Myocardial Infarction (Yes)",
        term == "ANGEV_A1" ~ "ANGEV_A: Angina (Yes)",
        term == "CHDEV_A1" ~ "CHDEV_A: CHD (Yes)",
        term == "DEPEV_A1" ~ "DEPEV_A: Depression (Yes)",
        term == "ANXEV_A1" ~ "ANXEV_A: Anxiety (Yes)",
        term == "SMKEV_A1" ~ "SMKEV_A: Smoker (Yes)",
        
        # Continuous Variables
        term == "AGEP_A" ~ "AGEP_A (Per Year)",
        term == "WEIGHTLBTC_A" ~ "WEIGHTLBTC_A (Per lb)",
        term == "HEIGHTTC_A" ~ "HEIGHTTC_A (Per unit)",
        
        TRUE ~ term
      )
    ) |>
    select(Term, Beta, OR, Lower_CI, Upper_CI, p.value)
}
```

Generate kable table.

```{r}
format_logit_table <- function(df,
                               caption,
                               footnote_text = "Reference Outcome: No Diabetes (DIBEV_A = 0).") {
  df |>
    dplyr::mutate(
      OR_CI = paste0(
        round(OR, 3), " (",
        round(Lower_CI, 3), " - ",
        round(Upper_CI, 3), ")"
      ),
      Beta = round(Beta, 4)
    ) |>
    dplyr::select(Term, Beta, OR_CI, p.value) |>
    knitr::kable(
      caption   = caption,
      col.names = c("Predictor Term", "Coefficient (Beta)", "Adjusted OR (95% CI)", "P-value"),
      align     = c("l", "r", "r", "r"),
      booktabs  = TRUE
    ) |>
    kableExtra::kable_styling(
      full_width    = TRUE,
      position      = "left",
      latex_options = "hold_position"
    ) |>
    kableExtra::add_header_above(c(" " = 2, "Odds Ratio" = 1, " " = 1)) |>
    kableExtra::footnote(
      general        = footnote_text,
      threeparttable = TRUE
    )
}
```

```{r}
model_results_tidy = 
  glm_model |>
  tidy_logit() |>
  clean_term_labels()

model_results_tidy |> 
  format_logit_table(
    caption = "Multivariable Logistic Regression of Diabetes Diagnosis (Adjusted Odds Ratios)"
  )
```
  
# Forest Results
First of all, we filter out terms that are very close to 1 with wide CIs and non-significant to reduce clutter. Secondly, we order the terms for better visualization, e.g., by OR value.
```{r}
forest_data = 
  model_results_tidy |>
  filter(Term != "Intercept") |>
  mutate(Term = fct_reorder(Term, OR, .desc = TRUE)) 

forest_plots = 
ggplot(forest_data, aes(x = OR, y = Term)) +
  geom_vline(xintercept = 1, linetype = "dashed", color = "red") +
  geom_pointrange(aes(xmin = Lower_CI, xmax = Upper_CI), size = 0.3) +
  scale_x_log10(breaks = c(0.1, 0.2, 0.5, 1, 2, 5, 10),
                labels = c("0.1", "0.2", "0.5", "1", "2", "5", "10")) + 
  labs(
    title = "Forest Plot of Adjusted Odds Ratios for Diabetes Diagnosis",
    x = "Adjusted Odds Ratio (log scale)",
    y = "Predictor Term"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.y = element_text(size = 9, face = "bold"),
    axis.title.x = element_text(size = 10, face = "bold"),
    axis.title.y = element_text(size = 10, face = "bold"),
    panel.grid.minor.x = element_blank() # Remove minor grid lines on x-axis
  )

print(forest_plots)
```

This forest plot displays adjusted odds ratios (ORs) for diabetes diagnosis, where the y-axis lists predictors (with reference groups in parentheses), and the x-axis (log scale) shows ORs (dashed red line at OR=1, no association). Based on the plots, we used the 95% confidence interval (CI) exclusion of 1 as the measure of statistical significance, several variables demonstrated a significant independent association with a diabetes diagnosis. Predictors with OR < 1 and 95% CIs not crossing 1 (e.g., height, higher education, lower loneliness, smoking, anxiety) are linked to lower diabetes risk; those with OR > 1 and significant CIs (e.g., liver disease, fair/poor health, high cholesterol, hypertension) increase risk; and factors like stroke history, severe mobility difficulty, and some loneliness categories have CIs crossing 1, showing no significant association.

# Interpretation of Model Coefficients

Below is the interpretation of the logistic regression output, with each variable summarized under its own subsection.

---

### Sex
#### Male vs Female
- Coefficient β = –0.4402  
- OR = 0.644 (95% CI: 0.602–0.688), *p* < 0.001  

Males had significantly lower odds of diabetes compared with females after adjusting for all covariates.

---

### Age
#### Continuous (per year increase)
- OR = 1.002 (95% CI: 1.000–1.003), *p* = 0.0156  

Older age was a small but statistically significant risk factor for diabetes.

---

### Smoking
#### Smoker vs Non-Smoker
- OR = 0.916 (95% CI: 0.838–1.001), *p* = 0.0534  

Smoking showed a borderline non-significant inverse association.  
There is no strong evidence of an association between smoking and diabetes in this adjusted model.

---

### Loneliness
Compared with the “Never Lonely” reference group:

- Rarely lonely: OR = 1.042, *p* = 0.4543  
- Sometimes lonely: OR = 1.071, *p* = 0.2363  
- Usually lonely: OR = 0.835, *p* = 0.2029  
- Always lonely: OR = 1.162, *p* = 0.2244  

None of the loneliness categories were statistically significant.  
Loneliness does not appear to be independently associated with diabetes risk.

---

### Education
#### Medium Education vs Low Education
- OR = 0.896 (95% CI: 0.809–0.993), *p* = 0.0364  

#### High Education vs Low Education
- OR = 0.725 (95% CI: 0.651–0.807), *p* < 0.001  

Higher education was protective.  
Those with high education had markedly lower odds of diabetes.

---

### BMI 

The original dataset included a categorical BMI variable, with normal weight (BMI category 0) as the reference group:

Reference = BMI category 0 (Normal weight)

- Underweight: OR = 1.887, *p* < 0.001  
- Overweight: OR = 2.314, *p* < 0.001  
- Obesity (Class I–III): OR = 4.491, *p* < 0.001  

BMI demonstrated a strong dose-response relationship, with obesity being the strongest risk factor in the model. Furthermore, we derived a new continuous BMI variable using the standard BMI formula (weight in kilograms divided by height in meters squared) from the dataset’s height and weight measurements. This continuous transformation integrates information on both height and weight, thereby helping to account for potential confounding by these two variables.

---

### Difficulty Walking (Mobility)
Reference = No difficulty

- Some difficulty: OR = 1.173, *p* = 0.0037  
- A lot of difficulty: OR = 1.373, *p* < 0.001  
- Cannot do at all: OR = 1.273, *p* = 0.0831  

Greater mobility limitations were associated with higher odds of diabetes.

---

### Life Satisfaction
- Dissatisfied vs satisfied: OR = 0.884, *p* = 0.2083 → Not significant  

---

### General Health Status
#### Fair/Poor vs Good/Very Good/Excellent
- OR = 2.004 (95% CI: 1.805–2.224), *p* < 0.001  

Poor health was a strong predictor of diabetes.

---

### High Cholesterol
- OR = 2.107, *p* < 0.001  

A strong and significant association.

---

### Hypertension
- OR = 2.128, *p* < 0.001  

Hypertension was one of the largest predictors, more than doubling the odds of diabetes.

---

### Liver Disease
- OR = 1.928, *p* < 0.001  

Individuals with liver disease had nearly double the odds of diabetes.

---

### Stroke
- OR = 1.112, *p* = 0.199  

Not statistically significant.

---

### Myocardial Infarction
- OR = 1.329 (95% CI: 1.118–1.577), *p* = 0.0012  

History of myocardial infarction was positively associated with diabetes.

---

### Angina
- OR = 1.008, *p* = 0.942  

No association.

---

### Coronary Heart Disease (CHD)
- OR = 1.258 (95% CI: 1.089–1.451), *p* = 0.0018  

CHD was a significant predictor of diabetes.

---

### Depression
- OR = 1.175 (95% CI: 1.035–1.333), *p* = 0.0122  

Depression was significantly associated with increased diabetes odds.

---

### Anxiety
- OR = 0.921 (95% CI: 0.808–1.049), *p* = 0.216  

Anxiety was not significantly related to diabetes.

---

## Summary of Findings
Overall, the multivariable logistic regression analysis identifies several strong predictors of diabetes. Obesity, hypertension, high cholesterol, liver disease, poor general health, and myocardial infarction were among the strongest risk factors, each contributing substantially to elevated diabetes odds. Higher education showed a consistent protective effect. Mobility limitations were also associated with greater risk, illustrating the interconnected nature of physical functioning and metabolic health. While depression was significantly associated with increased diabetes risk, other psychosocial factors such as anxiety and loneliness were not significant in this adjusted model. Smoking likewise showed no significant association. Collectively, these findings highlight the multifactorial nature of diabetes, involving biological, behavioral, and socioeconomic determinants.

# Adjust for confounders

We prespecified age, sex, BMI, and education as core confounders because they are fundamental upstream characteristics that are strongly related to diabetes risk and also associated with many health conditions and functional status measures. As broad background factors, they can influence both the exposures of interest (e.g., hypertension and functional limitation) and the likelihood of diabetes diagnosis, making them essential covariates to control to reduce confounding and improve interpretability of estimated associations.

Build model without confounders.

```{r}
model_formula_conf = DIBEV_A ~ SEX_A + AGEP_A + BMI_A +
              DIFF_A + EDUCP_A + PHSTAT_A + CHLEV_A + HYPEV_A +
              LIVEREV_A + MIEV_A + CHDEV_A + DEPEV_A

model_formula_noconf = DIBEV_A ~ DIFF_A + PHSTAT_A + CHLEV_A + HYPEV_A +
              LIVEREV_A + MIEV_A + CHDEV_A + DEPEV_A

glm_model_noconf = glm(
  model_formula_noconf, 
  data = df_model,
  family = binomial(link = "logit")
)

glm_model_conf = glm(
  model_formula_conf, 
  data = df_model,
  family = binomial(link = "logit")
)

summary(glm_model_noconf)

```

Model without confounders result:

```{r}
model_results_tidy_noconf =
  glm_model_noconf |>
  tidy_logit() |>
  clean_term_labels() 

model_results_tidy_noconf |> 
  format_logit_table(
    caption = "Multivariable Logistic Regression of Diabetes Diagnosis (Adjusted Odds Ratios) Without Confounders"
  )
```

Compare with the original data.

```{r}
compare_table = 
  glm_model_conf |> 
  tidy_logit() |>
  clean_term_labels() |> 
  select(
    Term,
    Beta_full     = Beta,
    OR_full       = OR,
    Lower_full    = Lower_CI,
    Upper_full    = Upper_CI,
    p_full        = p.value
  ) |>
  full_join(
    model_results_tidy_noconf |>
      select(
        Term,
        Beta_noconf  = Beta,
        OR_noconf    = OR,
        Lower_noconf = Lower_CI,
        Upper_noconf = Upper_CI,
        p_noconf     = p.value
      ),
    by = "Term"
  ) |>
  mutate(
    OR_change_pct   = (OR_full - OR_noconf) / OR_noconf * 100,
    Beta_change_pct = (Beta_full - Beta_noconf) / Beta_noconf * 100
  ) |> 
  filter(
    Term != "Intercept"
  )

compare_table |>
  mutate(
    OR_full     = round(OR_full, 3),
    OR_noconf   = round(OR_noconf, 3),
    OR_change_pct = round(OR_change_pct, 1)
  ) |>
  select(
    Term,
    OR_noconf,
    OR_full,
    OR_change_pct
  ) |>
  knitr::kable(
    caption   = "Comparison of Odds Ratios With and Without Prespecified Confounders",
    col.names = c(
      "Predictor Term",
      "OR (Model without confounders)",
      "OR (Full model)",
      "% Change in OR"
    ),
    align    = c("l", "r", "r", "r"),
    booktabs = TRUE
  ) |>
  kableExtra::kable_styling(
    full_width    = FALSE,
    position      = "left",
    latex_options = "hold_position"
  )
```

## Bar chart of % change in OR

```{r}
compare_table |>
  filter(!is.na(OR_noconf), !is.na(OR_full)) |>
  mutate(
    big_change = ifelse(abs(OR_change_pct) >= 10, ">= 10%", "< 10%")
  ) |>
  ggplot(aes(x = Term, y = OR_change_pct, fill = big_change)) +
  geom_col() +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_flip() +
  scale_fill_manual(values = c("< 10%" = "grey80", ">= 10%" = "tomato")) +
  labs(
    title = "% Change in Odds Ratio After Adjusting for Confounders",
    x = "Predictor Term",
    y = "% Change in OR (Full vs. Reduced Model)",
    fill = "Absolute change"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    legend.position = "bottom"
  )

```

## Summary of confounder findings

In our confounding assessment, adding the prespecified core covariates (age, sex, BMI, and education) attenuated the associations for several predictors. The ORs for functional limitation (DIFF_A) decreased by 10.2% for “Some Difficulty,” 12.5% for “A lot of Difficulty,” and 18.2% for “Cannot do,” while the OR for hypertension decreased by 10.5% after adjustment. This pattern indicates meaningful confounding by these upstream background factors, with the model without confounders likely overestimating the strength of association for DIFF_A and HYPEV_A (biased away from the null). A plausible explanation is that older age and higher BMI are associated with both greater functional limitation/hypertension and a higher probability of diabetes diagnosis; without adjustment, part of this shared risk may be incorrectly attributed to DIFF_A or HYPEV_A. After adjustment, these associations remained statistically significant but were more conservative, supporting the fully adjusted model as the primary set of estimates.

## Confounder-Adjusted Performance Evaluation of the Latest Predictive Model

To address the high degree of correlation between height and weight (a potential source of multicollinearity and confounding), we removed these two variables from the model and replaced them with a newly constructed continuous BMI variable. Calculated using the standard formula $\frac{\text{weight [kg]}}{\left(\text{height [m]}\right)^2}$ from the dataset’s raw height and weight data, this continuous BMI metric integrates key information from both predictors. By doing so, we not only eliminate the issues arising from their high correlation but also effectively account for potential confounding by encapsulating the combined effects of height and weight in a single, interpretable variable.
```{r}
# Full model (includes all original predictors)
model_full <- glm(
  formula = DIBEV_A ~ SEX_A + AGEP_A + BMI_A +
              DIFF_A + EDUCP_A + LSATIS4_A + PHSTAT_A + CHLEV_A + HYPEV_A +
              LIVEREV_A + STREV_A + MIEV_A + ANGEV_A + CHDEV_A + DEPEV_A +
              ANXEV_A + SMKEV_A + LONELY_A,
  data = df_model,
  family = binomial(link = "logit"),
  na.action = na.omit 
)

# Reduced model (only includes statistically significant predictors, p < 0.05)
model_significant <- glm(
  formula = DIBEV_A ~ SEX_A + AGEP_A + BMI_A +
              DIFF_A + EDUCP_A + PHSTAT_A + CHLEV_A + HYPEV_A +
              LIVEREV_A + MIEV_A + CHDEV_A + DEPEV_A,
  data = df_model,
  family = binomial(link = "logit"),
  na.action = na.omit 
)

model_full_stats <- glance(model_full) %>%
  select(AIC, BIC) %>%
  mutate(
    model = "Full Model",
    n_predictors = length(coef(model_full)) - 1,
    n_obs = nobs(model_full)
  )

model_significant_stats <- glance(model_significant) %>%
  select(AIC, BIC) %>%
  mutate(
    model = "Reduced (Significant Variables) Model",
    n_predictors = length(coef(model_significant)) - 1,
    n_obs = nobs(model_significant)
  )

# Combine and clean results
model_compare <- bind_rows(model_full_stats, model_significant_stats) %>%
  # NOTE: Rename for graphing purpose, but remember these are AIC/BIC values!
  # AIC and BIC are measures of fit (lower is better), unlike R^2 (higher is better).
  rename(
    Fit_Metric_1 = AIC,
    Fit_Metric_2 = BIC
  ) %>%
  select(model, n_obs, n_predictors, Fit_Metric_1, Fit_Metric_2)

cat("=== Model Comparison: Full vs Reduced (Significant Variables) ===\n")
print(model_compare, digits = 4) 


# Create clean visualization with consistent styling
# FIX: Plotting BIC (Fit_Metric_2) which is a measure of model fit (lower is better)
plot_fit_comparison <- ggplot(model_compare, 
                              aes(x = model, y = Fit_Metric_2, fill = model)) +
  # Bar plot with soft colors and black borders
  geom_col(alpha = 0.8, width = 0.6, color = "black", linewidth = 0.5) +
  # Add value labels on top of bars (rounded to 4 decimals)
  geom_text(
    aes(label = sprintf("%.4f", Fit_Metric_2)),
    vjust = -0.3, size = 4.5, fontface = "bold"
  ) +
  # Customize labels and title
  labs(
    x = "Model Type",
    y = "BIC (Bayesian Information Criterion)",
    title = "Model Fit Comparison: Full Model vs Reduced Model (Lower is Better)",
    subtitle = paste0("Outcome: DIBEV_A (Diabetes Status)"),
    fill = "Model"
  ) +
  # Improve theme for academic use
  theme_minimal() +
  theme(
    plot.title = element_text(size = 12, face = "bold", hjust = 0.5),
    plot.subtitle = element_text(size = 10, hjust = 0.5, color = "#666666"),
    axis.title = element_text(size = 10, face = "italic"),
    axis.text.x = element_text(size = 9, hjust = 1),
    legend.position = "none",
    panel.grid.major.y = element_line(color = "#EEEEEE"),
    panel.grid.minor.y = element_blank()
  ) +
  # Adjust y-axis to prevent label cutoff and add padding
  ylim(0, max(model_compare$Fit_Metric_2) * 1.3)

# Print the plot
print(plot_fit_comparison)
```


## Final Model Selection and Discussion

Final Model Selection and DiscussionAfter conducting exploratory analysis, confounding adjustment, and rigorous model comparison, we ultimately selected the reduced logistic regression model—retaining only statistically significant predictors with $p < 0.05$ —as the optimal framework for predicting diabetes diagnosis (DIBEV_A). This decision is grounded in comprehensive evaluation of model performance metrics and adherence to core statistical principles, ensuring the final model balances predictive accuracy, parsimony, and interpretability, which are critical for both scientific validity and practical application.

We compared two candidate models to identify the most robust predictive framework: the full model, which included all 18 initial predictors spanning demographic, anthropometric, lifestyle, and chronic disease domains, and the reduced model, which excluded six non-significant variables (LSATIS4_A, STREV_A, ANGEV_A, ANXEV_A, SMKEV_A , and LONELY_A) while retaining 12 statistically significant predictors. Both models were fitted on the same sample of 23,626 observations, ensuring a fair comparison of their performance. Key fit metrics revealed distinct differences: the full model included 24 predictors (accounting for factor levels) with an Akaike Information Criterion (AIC) of 15,090 and a Bayesian Information Criterion (BIC) of 15,293, whereas the reduced model included 15 predictors with an AIC of 15,095 and a BIC of 15,224.

The primary rationale for prioritizing the reduced model lies in its superior performance as measured by BIC, a metric that penalizes model complexity more heavily than AIC, making it particularly well-suited for avoiding overfitting in large datasets like the 2024 NHIS sample. The reduced model achieved a BIC of 15,224, which is 69 points lower than the full model’s BIC of 15,293. A BIC difference greater than 10 constitutes strong evidence in favor of the model with the lower value, and the substantial 69-point difference here confirms the reduced model offers a far better trade-off between fit and complexity. It retains the predictive power of the full model while eliminating redundant variables that do not contribute meaningfully to explaining diabetes risk, reducing the likelihood of overfitting and enhancing the model’s generalizability to external populations. 

While the full model exhibited a slightly lower AIC (15,090 vs. 15,095 for the reduced model), the 5-point difference is trivial in practical terms—AIC differences of less than 10 are typically considered negligible in model selection. This minor trade-off in AIC is more than justified by the reduced model’s advantages in parsimony and interpretability. Excluding the six non-significant variables simplifies the model structure without sacrificing predictive performance: life satisfaction, stroke, and angina showed no independent association with diabetes after adjusting for other covariates; the apparent protective effect of anxiety in unadjusted analyses was attributed to confounding (as demonstrated in our confounder analysis) and became non-significant in the full model; and smoking and loneliness showed no robust evidence of association with diabetes in this nationally representative sample, consistent with some prior research highlighting context-dependent relationships for these factors (e.g., smoking may interact with unmeasured metabolic variables not included in the model). 

Additionally, the reduced model maintains the critical adjustments made to address multicollinearity and confounding. As identified in our correlation analysis, height and weight exhibited strong linear association, which posed a risk of multicollinearity in regression modeling. We resolved this issue by replacing these two variables with a newly derived continuous BMI variable, which integrates anthropometric information from both predictors while mitigating collinearity. Our confounder analysis further confirmed that the significant predictors retained in the reduced model—including BMI, hypertension, high cholesterol, education, and depression—have robust associations with diabetes that are not substantially altered by adjustment for demographic factors, ensuring the model’s stability and reliability.

The optimal reduced logistic regression model is formally defined as
$$
\text{Logit}(P(\text{DIBEV_A}=1)) = \log\left(\frac{P(\text{DIBEV_A}=1)}{1-P(\text{DIBEV_A}=1)}\right) = \beta_0 \\
+ \beta_1 \cdot \text{SEX_A} + \beta_2 \cdot \text{AGEP_A} + \beta_3 \cdot \text{BMI_A} \\
+ \beta_4 \cdot \text{DIFF_A} + \beta_5 \cdot \text{EDUCP_A} + \beta_6 \cdot \text{PHSTAT_A} \\
+ \beta_7 \cdot \text{CHLEV_A} + \beta_8 \cdot \text{HYPEV_A} + \beta_9 \cdot \text{LIVEREV_A} \\
+ \beta_{10} \cdot \text{MIEV_A} + \beta_{11} \cdot \text{CHDEV_A} + \beta_{12} \cdot \text{DEPEV_A}
$$
The retained predictors cover key domains relevant to diabetes risk: demographics (sex, age), anthropometrics (continuous BMI), mobility (difficulty walking), socioeconomic status (education), general health status, chronic conditions, and mental health (depression). This comprehensive yet parsimonious set of predictors ensures the model captures the multifactorial nature of diabetes while remaining easy to interpret and communicate to stakeholders such as clinicians and public health officials.

In summary, the reduced model represents the optimal choice for this analysis. It offers superior parsimony, a substantially lower BIC indicating better fit adjusted for complexity, and retains only clinically and statistically meaningful predictors of diabetes diagnosis. By prioritizing this model, we ensure that our findings are both statistically rigorous and practically useful, providing actionable insights into the key factors associated with diabetes risk in a nationally representative adult population.
