---
title: "data"
output: 
  html_document:
    toc: true
    toc_float: true
---

```{r setup, include=FALSE}
library("gtsummary")
library("gt")
library("here")
```


```{r}
source(here("data", "datacleaning.R"))
```

# Statistical Description
```{r}
table1 =
  nhis24_df_fit |>
  tbl_summary(
    missing = "no",
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    type = list(
      all_categorical() ~ "categorical"
    )
  ) |>
  bold_labels() |>
  as_gt() |>
  gt::tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_title(groups = "title")
  ) |>
  gt::opt_table_outline() |>
  gt::opt_row_striping() |>
  gt::opt_align_table_header(align = "left")

table1
```
## Variable Descriptions

### SEX_A
#### Original Meaning
SEX_A identifies the biological sex of the respondent (1 = Male, 2 = Female) as defined in the NHIS adult codebook.

#### Recoding
The variable was recoded into a binary indicator where:

- 1 = Male  
- 0 = Female  

---

### AGEP_A
#### Original Meaning
AGEP_A represents the respondent’s age in years.

#### Recoding
No recoding was performed; the variable is used as a continuous numeric measure.

---

### DIBEV_A
#### Original Meaning
Indicates whether the respondent has ever been told by a doctor that they have diabetes (1 = Yes, 2 = No).

#### Recoding
The variable was converted to:

- 1 = Diabetes  
- 0 = No diabetes  
- Responses 7, 8, and 9 (refused/don’t know/not ascertained) were removed.

---

### BMICATD_A
#### Original Meaning
BMI category assigned based on measured height and weight:  
1 = Underweight  
2 = Healthy weight  
3 = Overweight  
4–6 = Obesity classes I–III  

#### Recoding
BMI was regrouped into four meaningful analytic categories:

- 0 = Healthy weight  
- 1 = Underweight  
- 2 = Overweight  
- 3 = Obesity (classes I–III combined)

---

### WEIGHTLBTC_A
#### Original Meaning
Measured body weight in pounds.

---

### HEIGHTTC_A
#### Original Meaning
Measured height in inches.

---

### DIFF_A
#### Original Meaning
Self-reported difficulty walking or climbing steps (1 = None, 2 = Some, 3 = A lot, 4 = Cannot do).

#### Recoding
Recoded as:

- 0 = No difficulty  
- 1 = Some difficulty  
- 2 = A lot of difficulty  
- 3 = Cannot do  

---

### EDUCP_A
#### Original Meaning
Highest educational attainment (ranging from no schooling to graduate degree).

#### Recoding
Grouped as:

- 0 = Low education  
- 1 = Medium education  
- 2 = High education  

---

### LSATIS4_A
#### Original Meaning
Life satisfaction rating (1 = Very satisfied → 4 = Very dissatisfied).

#### Recoding
Combined into:

- 0 = Satisfied (1–2)  
- 1 = Dissatisfied (3–4)

---

### PHSTAT_A
#### Original Meaning
General health status (Excellent / Very good / Good / Fair / Poor).

#### Recoding
Reclassified into:

- 0 = Good health (Excellent–Good)  
- 1 = Poor health (Fair/Poor)

---

### CHLEV_A
#### Original Meaning
History of high cholesterol (1 = Yes, 2 = No).

#### Recoding
Converted to:

- 1 = Yes  
- 0 = No  

---

### HYPEV_A
#### Original Meaning
History of hypertension.

#### Recoding

- 1 = Yes
- 0 = No.

---

### LIVEREV_A
#### Original Meaning
History of liver disease.

#### Recoding
- 1 = Yes
- 0 = No.

---

### STREV_A
#### Original Meaning
History of stroke.

#### Recoding
- 1 = Yes
- 0 = No.

---

### MIEV_A
#### Original Meaning
History of myocardial infarction (heart attack).

#### Recoding
- 1 = Yes
- 0 = No.

---

### ANGEV_A
#### Original Meaning
History of angina.

#### Recoding
- 1 = Yes
- 0 = No.

---

### CHDEV_A
#### Original Meaning
History of coronary heart disease.

#### Recoding
- 1 = Yes
- 0 = No.

---

### DEPEV_A
#### Original Meaning
Ever told had depression.

#### Recoding
- 1 = Yes
- 0 = No.

---

### ANXEV_A
#### Original Meaning
Ever told had anxiety.

#### Recoding
- 1 = Yes
- 0 = No.

---

### SMKEV_A
#### Original Meaning
Smoking history (1 = Yes, 2 = No).

#### Recoding
Binary coding:
- 1 = Yes (Ever smoked) 
- 0 = No (Never smoked) 

---

### LONELY_A
#### Original Meaning
Frequency of loneliness (1 = Always → 5 = Never).

#### Recoding
Converted to an ordered scale:

- 4 = Always lonely  
- 3 = Often  
- 2 = Sometimes  
- 1 = Rarely  
- 0 = Never  



# Correlation Analysis

To better understand the relationships between the variables in our dataset, we conducted a correlation analysis. The dataset was obtained from the 2024 NHIS (National Health Interview Survey), which contains information on demographic characteristics, health conditions, and behavioral factors for a nationally representative sample of adults. The cleaned dataset included variables such as age, sex, weight, height, and various health indicators.

Why we conducted correlation analysis:
Correlation analysis serves as a preliminary step to identify variables that are linearly associated. Understanding these relationships helps us:

Detect potential multicollinearity before performing regression analyses or more complex modeling.

Identify variables with meaningful associations that may warrant deeper investigation in subsequent analyses.

Provide evidence-based justification for including or excluding certain variables in predictive or explanatory models.

Method:
We computed pairwise Pearson correlations for all numeric and binary-coded variables. For numeric vs binary variables, the point-biserial correlation is equivalent to Pearson correlation. Correlations with an absolute value greater than 0.3 were considered moderate or strong associations. Additionally, diagonal elements (self-correlations) were excluded from interpretation, and symmetric duplicates were removed to ensure clarity.
```{r}
# Select numeric variables
numeric_vars <- nhis24_df_fit[, sapply(nhis24_df_fit, is.numeric)]

# Compute correlation matrix
cor_matrix <- cor(numeric_vars, use = "pairwise.complete.obs")

# Show matrix
print(cor_matrix)

heatmap(cor_matrix, symm = TRUE)
```

Key Findings:
From our analysis, the following moderate-to-strong correlations were observed:
```{r}
# > 0.03 in cor_matrix
# Convert correlation matrix to a data frame
cor_df <- as.data.frame(as.table(cor_matrix))

# Rename columns
colnames(cor_df) <- c("Var1", "Var2", "Correlation")
# Filter |correlation| > 0.3 but exclude self-correlations
high_cor <- cor_df |> 
  filter(abs(Correlation) > 0.3 & Var1 != Var2)

# remove duplicate
high_cor_unique <- high_cor %>%
  rowwise() %>%
  mutate(pair = paste(sort(c(Var1, Var2)), collapse = "_")) %>%
  distinct(pair, .keep_all = TRUE) %>%
  select(-pair) |> 
  arrange(desc(Correlation))

knitr::kable(high_cor_unique, caption = "Variables with Correlation > 0.3")
```

Importance and Academic Support:
Identifying these associations is crucial for informing future multivariate analyses, such as logistic regression or generalized linear models. For instance, strong correlations like height and weight or anxiety and depression suggest potential multicollinearity, which must be accounted for to prevent biased model estimates. Academic literature supports this approach; for example:

Dormann et al. (2013) highlight that assessing correlations is a standard procedure for detecting multicollinearity in ecological and health data analyses.

Studies on mental health populations have repeatedly observed strong comorbidity between anxiety and depression, consistent with our findings (Kessler et al., 2003).

In summary, this correlation analysis provides a foundation for selecting variables and interpreting relationships in subsequent modeling, ensuring that our future in-depth analyses are both statistically sound and aligned with existing evidence.

# Confounder Analysis



















