---
title: "Data"
---

```{r setup, include=FALSE}
library("gtsummary")
library("gt")
library("here")
library(ggplot2)
```


```{r}
source(here("data", "datacleaning.R"))
```

# Statistical Description
```{r}
table1 =
  nhis24_df_fit |>
  tbl_summary(
    missing = "no",
    statistic = list(
      all_continuous() ~ "{mean} ({sd})",
      all_categorical() ~ "{n} ({p}%)"
    ),
    type = list(
      all_categorical() ~ "categorical"
    )
  ) |>
  bold_labels() |>
  as_gt() |>
  gt::tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_title(groups = "title")
  ) |>
  gt::opt_table_outline() |>
  gt::opt_row_striping() |>
  gt::opt_align_table_header(align = "left")

table1
```
## Variable Descriptions

### SEX_A
#### Original Meaning
SEX_A identifies the biological sex of the respondent (1 = Male, 2 = Female) as defined in the NHIS adult codebook.

#### Recoding
The variable was recoded into a binary indicator where:

- 1 = Male  
- 0 = Female  

---

### AGEP_A
#### Original Meaning
AGEP_A represents the respondent’s age in years.

#### Recoding
No recoding was performed; the variable is used as a continuous numeric measure.

---

### DIBEV_A
#### Original Meaning
Indicates whether the respondent has ever been told by a doctor that they have diabetes (1 = Yes, 2 = No).

#### Recoding
The variable was converted to:

- 1 = Diabetes  
- 0 = No diabetes  
- Responses 7, 8, and 9 (refused/don’t know/not ascertained) were removed.

---

### BMICATD_A & BMI_A
#### Original Meaning
The original dataset:
BMI category assigned based on measured height and weight:  
1 = Underweight  
2 = Healthy weight  
3 = Overweight  
4–6 = Obesity classes I–III  

#### Recoding
BMI was regrouped into four meaningful analytic categories:

- 0 = Healthy weight  
- 1 = Underweight  
- 2 = Overweight  
- 3 = Obesity (classes I–III combined)

#### BMI_A
We constructed a new continuous BMI variable using the standard BMI formula (weight in kilograms divided by height in meters squared). 

New variable `BMI_A` is a continuous variable that integrates information on both height and weight, thereby helping to account for potential confounding by these two variables.

---

### WEIGHTLBTC_A
#### Original Meaning
Measured body weight in pounds.

---

### HEIGHTTC_A
#### Original Meaning
Measured height in inches.

---

### DIFF_A
#### Original Meaning
Self-reported difficulty walking or climbing steps (1 = None, 2 = Some, 3 = A lot, 4 = Cannot do).

#### Recoding
Recoded as:

- 0 = No difficulty  
- 1 = Some difficulty  
- 2 = A lot of difficulty  
- 3 = Cannot do  

---

### EDUCP_A
#### Original Meaning
Highest educational attainment (ranging from no schooling to graduate degree).

#### Recoding
Grouped as:

- 0 = Low education  
- 1 = Medium education  
- 2 = High education  

---

### LSATIS4_A
#### Original Meaning
Life satisfaction rating (1 = Very satisfied → 4 = Very dissatisfied).

#### Recoding
Combined into:

- 0 = Satisfied (1–2)  
- 1 = Dissatisfied (3–4)

---

### PHSTAT_A
#### Original Meaning
General health status (Excellent / Very good / Good / Fair / Poor).

#### Recoding
Reclassified into:

- 0 = Good health (Excellent–Good)  
- 1 = Poor health (Fair/Poor)

---

### CHLEV_A
#### Original Meaning
History of high cholesterol (1 = Yes, 2 = No).

#### Recoding
Converted to:

- 1 = Yes  
- 0 = No  

---

### HYPEV_A
#### Original Meaning
History of hypertension.

#### Recoding

- 1 = Yes
- 0 = No.

---

### LIVEREV_A
#### Original Meaning
History of liver disease.

#### Recoding
- 1 = Yes
- 0 = No.

---

### STREV_A
#### Original Meaning
History of stroke.

#### Recoding
- 1 = Yes
- 0 = No.

---

### MIEV_A
#### Original Meaning
History of myocardial infarction (heart attack).

#### Recoding
- 1 = Yes
- 0 = No.

---

### ANGEV_A
#### Original Meaning
History of angina.

#### Recoding
- 1 = Yes
- 0 = No.

---

### CHDEV_A
#### Original Meaning
History of coronary heart disease.

#### Recoding
- 1 = Yes
- 0 = No.

---

### DEPEV_A
#### Original Meaning
Ever told had depression.

#### Recoding
- 1 = Yes
- 0 = No.

---

### ANXEV_A
#### Original Meaning
Ever told had anxiety.

#### Recoding
- 1 = Yes
- 0 = No.

---

### SMKEV_A
#### Original Meaning
Smoking history (1 = Yes, 2 = No).

#### Recoding
Binary coding:
- 1 = Yes (Ever smoked) 
- 0 = No (Never smoked) 

---

### LONELY_A
#### Original Meaning
Frequency of loneliness (1 = Always → 5 = Never).

#### Recoding
Converted to an ordered scale:

- 4 = Always lonely  
- 3 = Often  
- 2 = Sometimes  
- 1 = Rarely  
- 0 = Never  






