---
title: "Proposal"
---
```{r echo=FALSE}
library(knitr)
library(dplyr)
```

# Group Members:

Yaxuan Zheng (yz5248)

Yirou Hu (yh3964)

Ruihan Ding (rd3227)

Su Yan (sy3352)

Jinyao Jiao (jj3564)

# Tentative Project Title

Cross-Sectional Analysis of Different Covariates and Type 2 Diabetes Risk: A Generalized Linear Model Approach

# Purpose and Importance of T2D Analysis

Type 2 diabetes (T2D) remains a critical public health challenge worldwide. While age and BMI are primary risk factors, psychological factors such as anxiety, depression, and loneliness, alongside behavioral factors like smoking, are increasingly recognized as meaningful predictors of T2D prevalence.

Our project focuses on quantifying the strength and significance (coefficients) of the association between these non-traditional psychosocial and lifestyle factors and the prevalence of T2D, utilizing the robust Generalized Linear Model (Logistic Regression) framework. This analysis will provide detailed description of coefficient, offering crucial insights for public health interventions targeting T2D prevention.

# Variables

## Primary Outcome Table
```{r}
primary_outcome_df <- data.frame(
  Variable = c("`DIBEV_A`"),
  Description = c("Ever had diabetes"),
  Type = c("Dichotomous"),
  Coding_in_Analysis = c("**1 (Yes) vs 0 (No)**")
)

kable(
  primary_outcome_df,
  col.names = c("Variable", "Description", "Type", "Coding in Analysis"),
  caption = "Primary Outcome Variable",
  align = "l"
)
```

## Covariates Table
```{r}

covariates_df <- data.frame(
  Variable = c(
    "`SEX_A`", "`AGEP_A`", "`REGION`", "`BMICATD_A`", "`WEIGHTLBTC_A`", 
    "`HEIGHTTC_A`", "`DIFF_A`", "`DIBAGETC_A`", "`DIBTYPE_A`", "`EDUCP_A`", 
    "`LSATIS4_A`", "`PHSTAT_A`", "`CHLEV_A`", "`HYPEV_A`", "`LIVEREV_A`", 
    "`STREV_A`", "`MIEV_A`", "`ANGEV_A`", "`CHDEV_A`", "`DEPEV_A`", 
    "`ANXEV_A`", "`SMKEV_A`", "`LONELY_A`"
  ),
  Description = c(
    "Sex", "Age (Top-coded)", "Geographical Region", "Categorical BMI", "Weight (lbs, Top-coded)", 
    "Height (in, Top-coded)", "Difficulty walking/climbing", "Age first diagnosed (Type 2)", 
    "Type of diabetes", "Educational level", "Life satisfaction", "General health status (Self-rated)", 
    "Ever had high cholesterol", "Ever had hypertension", "Ever had liver condition/cirrhosis", 
    "Ever had a stroke", "Ever had a heart attack (MI)", "Ever had angina", 
    "Ever had coronary heart disease", "Ever had depression", "Ever had anxiety disorder", 
    "Ever smoked 100 cigarettes", "Frequency of feeling lonely"
  ),
  Category = c(
    "Demographic", "Demographic", "Demographic", "Anthropometric", "Anthropometric", 
    "Anthropometric", "Functional Status", "Diabetes Detail", "Diabetes Detail", "Socioeconomic", 
    "Psychosocial", "General Health", "Comorbidity", "Comorbidity", "Comorbidity", 
    "Comorbidity", "Comorbidity", "Comorbidity", "Comorbidity", "Psychosocial", 
    "Psychosocial", "Lifestyle", "Psychosocial"
  ),
  Type = c(
    "Dichotomous", "Continuous", "Categorical", "Ordinal", "Continuous", 
    "Continuous", "Ordinal", "Continuous", "Categorical", "Ordinal", 
    "Ordinal", "Ordinal", "Dichotomous", "Dichotomous", "Dichotomous", 
    "Dichotomous", "Dichotomous", "Dichotomous", "Dichotomous", "Dichotomous", 
    "Dichotomous", "Dichotomous", "Ordinal"
  )
)

kable(
  covariates_df,
  col.names = c("Variable", "Description", "Category", "Type"),
  caption = "Comprehensive Covariates",
  align = "l"
)

```


# Intended Final Products

- Writting report capturing data cleaning, exploratory analysis (e.g. Rank the covariates according to how strongly they relate to T2D), modeling, visualization(distribution plots, KM curves, etc.) and discussion, structured in R Markdown and published in the team’s GitHub repository to detailedly explain our propose and results.

- Creating website summarizing motivation, methods, results, and key visual-insights directed to a broad audience.

- A short (~2-minute) screencast embedded in the website walking through the major analyses and findings.

# Anticipated Data Sources

This study utilizes data from the United States National Health Interview Survey (NHIS) in 2024, a national survey that monitors the health of the U.S. population. NHIS statistics provide insights into the health of people of different ages, sexes, races and ethnicities, geographic locations, and social and economic circumstances, making it a valuable resource for studying complex diseases such as T2D.

# Planned Analyses / Visualizations / Coding Challenges

## Planned Analyses:

- Kaplan–Meier survival curves stratified by levels of mental‐health and smoking risk factors.

- Single and Multivariable Cox proportional hazards models estimating hazard ratios for T2D onset associated with anxiety, depression and smoking, controlling for age, sex, BMI, socioeconomic status.

- Assess potential interactive relationships between key factors (including mental health indicators, smoking, and covariates like age, BMI, or socioeconomic status) to identify which pairs exhibit stronger synergistic or antagonistic effects on T2D risk.

- Exploratory Spatial Data Analysis

## Visualizations: 
Variable correlation hot map; KM curves; Forest plots of hazard ratios; Sex-strafitication; Age-Stratification; Variable interaction effect table; Survival curve; Scatter plot of EDA

## Coding challenges: 
We may face challenges such as harmonizing variable definitions between earlier and updated datasets, dealing with missing data (through imputation or sensitivity analyses), automating reproducible workflows using git and R Markdown, and applying various models to explore the relationship between covariates and T2D.

# Timeline
```{r, echo=FALSE}
knitr::kable(
  data.frame(
    Date = c("Nov 3 - 7", "Nov 10-14", "Nov 17 - 21", "Nov 24 - 28", "Dec 1 - 5", "Dec 6", "Dec 11"),
    Task = c(
      "Brainstorm and draft proposal",
      "Assign tasks & initial exploratory manipulation on data",
      "Analysis & Coding",
      "Analysis & Coding",
      "Report & Webpage & Screencast",
      "Peer assessment",
      "Prepare for in class discussion"
    ),
    Due = c(
      "Written proposal document",
      "Project review meeting with teaching team",
      "NA",
      "NA",
      "NA",
      "Finished project",
      "In class discussion of projects"
    )
  ),
  caption = "Project Timeline",
  align = "l"
)
```

