---
title: "Final Project Proposal"
author: "Yirou Hu, Yaxuan Zheng, Ruihan Ding, Su Yan, Jinyao Jiao "
date: "2025-11-05"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Group Members:

Yaxuan Zheng (yz5248)

Yirou Hu (yh3964)

Ruihan Ding (rd3227)

Su Yan (sy3352)

Jinyao Jiao (jj3564)

# Tentative Project Title

Cohort Analysis of Mental Health, Smoking and Type 2 Diabetes Risk: A Cox Survival Study

# Motivation for This Project

Type 2 diabetes (T2D) remains a critical public health challenge worldwide. Recent research indicates that psychological factors such as anxiety and depression, and behavioral factors like smoking, are meaningful predictors of T2D incidence. 

# Intended Final Products

- Writting report capturing data cleaning, exploratory analysis (e.g. Rank the covariates according to how strongly they relate to T2D), modeling, visualization(distribution plots, KM curves, etc.) and discussion, structured in R Markdown and published in the team’s GitHub repository to detailedly explain our propose and results.

- Creating website summarizing motivation, methods, results, and key visual-insights directed to a broad audience.

- A short (~2-minute) screencast embedded in the website walking through the major analyses and findings.

# Anticipated Data Sources

This study utilizes data from the United States National Health Interview Survey (NHIS) in 2024, a national survey that monitors the health of the U.S. population. NHIS statistics provide insights into the health of people of different ages, sexes, races and ethnicities, geographic locations, and social and economic circumstances, making it a valuable resource for studying complex diseases such as T2D.

# Planned Analyses / Visualizations / Coding Challenges

## Planned Analyses:

- Kaplan–Meier survival curves stratified by levels of mental‐health and smoking risk factors.

- Single and Multivariable Cox proportional hazards models estimating hazard ratios for T2D onset associated with anxiety, depression and smoking, controlling for age, sex, BMI, socioeconomic status.

- Assess potential interactive relationships between key factors (including mental health indicators, smoking, and covariates like age, BMI, or socioeconomic status) to identify which pairs exhibit stronger synergistic or antagonistic effects on T2D risk.

- Exploratory Spatial Data Analysis

## Visualizations: 
Variable correlation hot map; KM curves; Forest plots of hazard ratios; Sex-strafitication; Age-Stratification; Variable interaction effect table; Survival curve; Scatter plot of EDA

## Coding challenges: 
We may face challenges such as harmonizing variable definitions between earlier and updated datasets, dealing with missing data (through imputation or sensitivity analyses), automating reproducible workflows using git and R Markdown, and applying various models to explore the relationship between covariates and T2D.

# Timeline
```{r, echo=FALSE}
knitr::kable(
  data.frame(
    Date = c("Nov 3 - 7", "Nov 10-14", "Nov 17 - 21", "Nov 24 - 28", "Dec 1 - 5", "Dec 6", "Dec 11"),
    Task = c(
      "Brainstorm and draft proposal",
      "Assign tasks & initial exploratory manipulation on data",
      "Analysis & Coding",
      "Analysis & Coding",
      "Report & Webpage & Screencast",
      "Peer assessment",
      "Prepare for in class discussion"
    ),
    Due = c(
      "Written proposal document",
      "Project review meeting with teaching team",
      "NA",
      "NA",
      "NA",
      "Finished project",
      "In class discussion of projects"
    )
  ),
  caption = "Project Timeline",
  align = "l"
)
```

